Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-10-17 Appendix.Rmd
First commit

---
title: "Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

@@

\
**Corresponding author**: 
Rutuja Chitra-Tarak
\
E: arutuj@gmail.com
\
T: +1(505) 665 1647

# Summary

- Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching critical safety thresholds. Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled effective rooting depth (ERD, as the average depth of water extraction), for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and to leaf hydraulic vulnerability curves. We validated ERD against existing tree and soil stable water isotope data. 
- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
- The role of deep-water access in alleviating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.

\
Keywords: below-ground traits, deep-water access, drought-induced mortality, drought tolerance, rooting depths, tropical forest, life-history strategies

# Introduction
\
    Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al. 2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events. Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\
    Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance. Deep water access is arguably the most long-standing, yet under-studied mechanism. This mechanism assumes that plant available water varies with depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019]. Consequently, trees in the same forest may differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth. However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020]. DNA barcoding of roots [@Jones.2011] may be used to to estimate species-specific rooting depths or rooting profiles, although extent of sampling required may be cost-prohibitive and DNA barcode libraries for tropical forests are still under development. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

\
    Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region. We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.

\
    We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig. \@ref(fig:schematic)). This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for 8 species and developing trait-based proxies for the rest. We evaluated ERD against existing data on plant and soil water isotopes and tested coordination of ERD with hydraulic traits and mortality rates. Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

Fig. \@ref(fig:schematic) here.
Table \@ref(tab:hypo) here.

# Materials and Methods
## Study site description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014]. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].

\
In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture. The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

## Model ELM-FATES for soil matric potentials

### Model description
\
To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes. In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) [@Bonan:2013]. FATES is a community-based, open-source model used for studying climate-vegetation interactions. FATES is a vegetation demography model, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography [@Moorcroft.2001] approach. FATES couples to ELM by a common interface of water and carbon fluxes. Detailed descriptions for ELM and FATES can be found elsewhere [@Fisher.2010; @Fisher.2015; @Koven:2020].

\
ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, i}$ at depth z~i~ and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and conservation of these terms is expressed as follows:

\begin{equation}
\begin{split}
\Delta W_{can} + \Delta W_{sfc} + \sum_{i=1}^{15} \Delta W_{liq,i} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
\end{split}
(\#eq:water-balance)
\end{equation}

where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step (typically, 1 hour). The soil profile is discretized into 15 exponentially distributed soil layers with depth z $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).

### Model parameterization
\
To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\Psi_{soil}$ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

\begin{equation}
\Psi_{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\Psi_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\Psi_e$, $\theta_{s}$ and $b$ taht best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

\
We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode. As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

### Model calibration
\
We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), (ii) stream discharge from the Conrad catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m (1299 samples in total covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ refer to plot-wide averages).

For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states. We describe the objective function $O$ as :

\begin{align*}
\overline{O} = \frac{1}{6}\sum \left(RMSE^{*}(ET,~E_{v} + E_{g}) + RMSE^{*}(q_{run},~q_{over} + q_{drai}) + RMSE^{*}(\theta_{0-0.15m}, {W_{liq,~\overline{z}}})\\
+ RMSE^{*}(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + RMSE^{*}(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
(\#eq:obj-func)
\end{align*}

based on monthly sums for both $ET$ and $q_{run}$ and daily averages for $\theta$s. $W_{liq,~\overline{z}}$ is the average of daily $W_{liq,~z}$ for $z$ $\in$ {`r paste(soil.depths[1:4], collapse=", ")`} m. Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.

Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit with soil moisture dynamics at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1m ($RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

### Model evaluation
\
We evaluated ELM-FATES by calculating RMSE between simulated and observed long-term daily VWC for the depths of 0.1, 0.4 and 1 m (2016-2018), baesd on a dataset we had left-out during calibration. We obtained these observations from three horizontal time domain reflectometry probes at the depths of 0.1, 0.4 and 1 m at a point location near the eddy flux tower (see [SI](#SIData.VWC)).

\
We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up.

## Model for effective rooting depth (ERD)
\
To estimate effective rooting depth, we tested alternate models for growth (Eqs. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. In this study, we used maximum diurnal leaf conductance ($K_{leaf}$) approximated using rhizosphere soil water potential ($\Psi_{soil}$) as a proxy for the soil water limiting factor on growth. In our models $K_{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential. For each species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,~t~|~z}$, in the census interval $t$ for depth $z$ and for a given hydrological realisation $h$ is described as follows:

\begin{equation}
G_{s,~t~|~z,~h} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{leaf,~s,~i~|~z,~h} + VPD^{*}_{i}) LAI^{*}_{s,~doy~|~i}) + \epsilon_{t~|~z,~h}
(\#eq:M-chosen)
\end{equation}

where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term. See SI for equations for all the alternate models tested.

As VPD may have a non-linear relationship with growth, with growth increasing for the lower range of VPD, and decreasing for the higher range of VPD due to stomatal closure (reference), we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq. \@ref(eq:M-chosen).

For each species, we estimated diurnal maximum value of $K_{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi_{soil,~i~|~z,~h}$ in the leaf vulnerability curves defined below in Eq. \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:

\begin{equation}
K_{leaf,~s,~i~|~z,~h} = A_{s}e^{-B_{s}\Psi_{soil,~i~|~z,~h}}
(\#eq:k-soil)
\end{equation}

By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit) [@Chonggang, we could have retained multiple depths with non-differentiable AICs], and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :

\begin{equation}
ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(AIC_{s~|~z,~h} = \underset{\forall \, z \,\in\,Z}{min}(AIC_{s~|~z,~h})\Big) \cap \Big(r_{s~|~z,~h} \ge 0.71\Big)\Big\}
(\#eq:erd)
\end{equation} 
where $H$ $\in$ {1, .., 100}, refer to the 100 hydrological realisations and $Z$ $\in$ {`r paste(soil.depths, collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision.

### Model parameterization

#### Growth data
\
For growth estimates, to minimize the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, the diameter at breast height) and also species whose maximum height was $\ge$ 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun. We selected trees for which the height of the diameter measurements were unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015). For each individual tree (*i*), we calculated absolute growth rates ($G_{i,t}$; cm yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. We eliminated positive outliers as any growth rates amounting to > 75 mm yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017]. Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh. Any stem whose later dbh measurement fell < 4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017]. As tree growth rates vary with size [@Muller-Landau:2006], to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = G_{i,t} - f(d_{i, t})$). The diameter model of growth ($f(d_{i, t})$), a B-spline based polynomial equation with 5 degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig. S\@ref(fig:dbh-model)). We retained only those trees that had $Resid_{i,t}$ estimates for all census intervals (total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $Resid_{i,t}$. We then obtained `r length(erd.sp)` species-level time-series of medians of $Resid_{i,t}$ ($G_{t}$). These were obtained for only those species with complete records on at least three trees (median *n* = `r median(g.n$n)`, max *n* = `r max(g.n$n)` trees per species).

#### Leaf hydraulic vulnerability curves
\
We obtained leaf hydraulic vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al. (to be submitted; see SI for brief description of methods) described as:

\begin{equation}
K_{leaf,~s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + error
(\#eq:vc-exp)
\end{equation}
where $A$ and $B$ are fitted species-specific parameters. These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation. For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies. We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits, namely, WSG, the wood specific gravity after drying at 100^$\\circ$^C and LMA, the leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules for leaves receiving direct sunlight (@Wright:2010; see SI for details). We fitted polynomial equations described as,

\begin{equation}
A_{s} = `r acf[1]` `r acf[2]`B_{s} `r acf[3]`B_{s}^2 + `r acf[4]`B_{s}LMA_{s} + `r acf[5]`LMA_{s} `r acf.6`LMA_{s}^2 + error
(\#eq:a-mod)
\end{equation}

\begin{equation}
B_{s} = `r bcf[1]` `r bcf[2]`WSG_{s} + `r bcf[3]`WSG_{s}^2 + `r bcf[4]`WSG_{s}LMA_{s} + `r bcf[5]`LMA_{s} `r bcf[6]`LMA_{s}^2 + error
(\#eq:b-mod)
\end{equation}

As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq. \#ref(eq:vc-exp). For all the 31 ERD species we estimated species $\Psi_{50,~leaf}$ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).

Table \@ref(tab:symbols) here.

#### Leaf Area Index (LAI)
\
To explore the effect of species seasonality in leaf phenology on growth, we assumed species-specific seasonal curves for leaf area index ($LAI$, range 0-1, unitless). For evergreen species we assumed $LAI$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LAI$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LAI$ for the rest of the year as follows. We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless). Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LAI$ = 1 beginning DOY = 300 and obtained a $LAI$ by DOY curve (Fig. S\@ref(fig:leaf-cover)). We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used K standardized between 0 and 1 within species ($K^*$) in Eq. \@ref(eq:M-chosen).

#### Vapor Pressure Deficit (VPD)
\
We fitted a polynomial curve between mean midday values (11:00 to 15:00 hrs) of gross primary productivity (gC m^2^ day^-1^) and VPD (kPa) observations we collected from the eddy flux tower located 75 m from the 50-ha plot (for any day *i*, $GPP_i = `r vpd.cf[1]` + `r vpd.cf[2]`VPD_i + `r vpd.cf[3]`VPD_i^2 + error$; 2012-2017; Fig. S\@ref(fig:gpp-vpd)) and used $GPP$ standardized between 0 and 1 as $VPD^*$ in Eq. \@ref(eq:M-chosen)).

### Model validation and selection
\
To validate the ERD models we used data for xylem sap $\delta^2H$ (permil) as an independent observation of ERD. For each model of ERD (Eq. \@ref(eq:M-chosen) and Eqs. in SI), estimated species ERD was regressed against species xylem sap $\delta^2H$ (permil). We used xylem sap $\delta^2H$ values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil (Fig. S\@ref(fig:iso-data)). We removed species that are likely to have no leaves in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves. We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile. We were thus left with 7 species with data on $\delta^2H$ as well as modelled estimates of ERD.

## ERD coordination with above-ground hydraulic traits
\
To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$). See [SI](#SIData.brett) for data collection and estimation of these variables.

## Mortality analyses and species-specific drought experience
\
We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 35-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50 ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~. For each species we calculated mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$.

    We regressed species ERD against species ($Mort_{mean}$). To assess whether deeper ERD species escape the experience of significant soil drought, we calculated for each speceis the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi_{soil,~z = ERD}$) were below species $\Psi_{crit}$. $\Psi_{crit}$ was defined as $\Psi_{50,~leaf}$.

\
All statistical analyses were conducted in R statistical environment (version 3.5.3, @R:2019).

# Results
## Soil water dynamics by depth
\
The ELM-FATES simulations with the chosen best-fit parameter ensembles (Table \@ref(tab:param)) successfully captured the dynamics of soil moisture by depth (Fig. \@ref(fig:hydro-calib) A & Fig. \@ref(fig:hydro-calib)B), including the observations from the horizontal TDR probes that were not included in the calibration. These simulations also captured the dynamics in stream discharge (Fig. \@ref(fig:hydro-calib)C) and to a moderate extent the dynamics in evapotranspiration (Fig. \@ref(fig:hydro-calib)D). The reduction in parameter range in the best-fit ensembles compared to the global ranges tested showed that the above calibration was sensitive to the following parameters: the Ball-Berry stomatal slope parameter, the ELM rooting distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and our adjustment factor to soil hydraulic conductivity to account for macroporosity and direct flow paths (Table \@ref(tab:param)).

Fig. \@ref(fig:hydro-calib) here.

  Soil Water Potential dynamics ($\Psi_{soil,~z}$) by depth (z = 0.1 through 13) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig. S\@ref(fig:swp-droughts)). Nevertheless, across the simulation period (1990-2015) $\Psi_{soil,~z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig. S\@ref(fig:swp-all-depths)).

Fig. \@ref(fig:swp-droughts) here.

>>>>

---
title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
subtitle: 'Appendix'
author: ''
date: ''
output:
  bookdown::html_document2:
    bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  number_sections: no
theme: flatly
bookdown::pdf_document2: default
csl: new-phytologist.csl
bibliography: references.bib
number_sections: true
# header-includes: 
#   \usepackage{float} \floatplacement{figure}{H} 
#   \setcounter{table}{0}  
#   \renewcommand{\thetable}{S\arabic{table}} 
#   \setcounter{figure}{0} 
#   \renewcommand{\thefigure}{S\arabic{figure}}
---

@@

\documentclass{article}

\usepackage{chngcntr}
\begin{document}

\section{Introduction}

\subsection{Problem description}

\begin{figure}[htp]
  \centering Figure
  \caption{Call this figure 1.}
  \label{fig1}
\end{figure}

S\@ref(fig:other-erd-mods)). 
S\@ref(fig:erd-mort-census)). 
\appendix
\counterwithin{figure}{section}

\section{Figures}

\beginsupplement

```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2H$ for each of the ERD models tested (panels). For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $\\rho$ > 0 and *R*^2^ > 0.5 during ERD parameterisation. The number of species with an ERD estimate as well as $\\delta^2H$. or each model are shown in inset, along with the goodness-of-fit (*R*^2^) and *P*-values for a linear relationship between ERD and $\\delta^2H$. See Eqs. \\@ref(eq:M1)-\\@ref(eq:M6). See Table \\@ref(tab:ab-param) for the full form of scientific names.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
```
